# Git and GitHub

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE)
```

## Git
[Git](https://git-scm.com/) is a piece of version control software written by Linus Torvalds, the creator of Linux. It essentially tracks changes to local files.

Let's say that you've written some R code that visualises the total fertility rate in Nigeria. You've saved the R script (`script.R`) in a folder called 'vital_statistics' and the dataset (`fertility_rate.csv`) in a new subfolder called 'data'. The resulting plot is stored in the 'outputs' folder.

```
.
└── vital_statistics
    ├── script.R
    ├── data
    │   ├── fertility_rate.csv
    ├── outputs
```

::: grid
::: g-col-6
```{r, echo=TRUE, eval=FALSE}
library(tidyverse) ; library(ggtext)

df <- read_csv("data/fertility_rate.csv") |> 
  mutate(year = ymd(year, truncated = 2L))

ggplot(df, aes(x = year, y = value)) + 
  geom_line(colour = "#CC79A7", linewidth = 1) + 
  scale_x_date(expand = expansion(),
               breaks = seq.Date(as.Date("1960-01-01"), as.Date("2100-01-01"), by = "20 years"),
               date_labels = "%Y") + 
  scale_y_continuous(limits = c(0,10),
                     labels = scales::label_number(accuracy = 1, scale_cut = scales::cut_short_scale())) +
  labs(x = NULL, y = "Births per woman",
       title = "Total fertility rate",
       subtitle = "<span style = 'color:#757575;'>Nigeria, 1953 to 2100</span>",
       caption = "Source: U.S. Census Bureau, International Database") +
  theme_minimal(base_size = 14) +
  theme(
    plot.margin = unit(rep(1, 4), "cm"),
    panel.grid.major.y = element_blank(),
    panel.grid.minor = element_blank(),
    axis.line.x = element_line(colour = "#000000"),
    plot.title = element_text(size = rel(1.2), face = "bold"),
    plot.subtitle = element_markdown(margin = margin(b = 15)),
    plot.caption = element_text(size = rel(0.8), colour = "#707071", margin = margin(t = 20)),
    axis.title.y = element_text(size = rel(0.8), hjust = 0, margin = margin(r = 10))
  ) 

ggsave("outputs/plot.jpeg", dpi = 300, scale = 1)
```
:::

::: g-col-6
```{r}
library(tidyverse) ; library(ggtext)

df <- read_csv("data/fertility_rate.csv") |> 
  filter(name == "Nigeria") |> 
  mutate(year = ymd(year, truncated = 2L))

ggplot(df, aes(x = year, y = value)) + 
  geom_line(colour = "#CC79A7", linewidth = 1) + 
  scale_x_date(expand = expansion(),
               breaks = seq.Date(as.Date("1960-01-01"), as.Date("2100-01-01"), by = "20 years"),
               date_labels = "%Y") + 
  scale_y_continuous(limits = c(0,10),
                     labels = scales::label_number(accuracy = 1, scale_cut = scales::cut_short_scale())) +
  labs(x = NULL, y = "Births per woman",
       title = "Total fertility rate",
       subtitle = "<span style = 'color:#757575;'>Nigeria, 1953 to 2100</span>",
       caption = "Source: U.S. Census Bureau, International Database") +
  theme_minimal(base_size = 14) +
  theme(
    plot.margin = unit(rep(1, 4), "cm"),
    panel.grid.major.y = element_blank(),
    panel.grid.minor = element_blank(),
    axis.line.x = element_line(colour = "#000000"),
    plot.title = element_text(size = rel(1.2), face = "bold"),
    plot.subtitle = element_markdown(margin = margin(b = 15)),
    plot.caption = element_text(size = rel(0.8), colour = "#707071", margin = margin(t = 20)),
    axis.title.y = element_text(size = rel(0.8), hjust = 0, margin = margin(r = 10))
  )
```
:::
:::

However, you've been told to compare Nigeria's fertility rate with Rwanda's. This is where git comes in. Rather than creating a new file or overwriting it you can just track your file changes with Git.

1. Navigate to the Terminal tab in the Console pane of RStudio. You can also use the Command Prompt or PowerShell in Windows or the Terminal on a Mac.
2. Go to the folder containing your code and data: `cd /path/to/your/folder`{.bash}
3. Initialise the folder as a git repository or **repo**: `git init`{.bash}.
4. Check which files are currently untracked: `git status`{.bash}

```sh
$ git status
No commits yet

Untracked files:
  (use "git add <file>..." to include in what will be committed)
        data/
        outputs/
        script.R

nothing added to commit but untracked files present (use "git add" to track)
```

We have two empty folders and one file (`script.R`) that are currently untracked.

5. Put the files in the staging area: `git add -A`{.bash}
6. Then take a snapshot with `git commit`{.bash}

::: {.callout-tip}
## Git commit messages

It is good practice to add a descriptive message to each commit. For example, we might add `git commit -m "Nigeria fertility rate"`{.bash}. When we look back through the history of our commits we can then see at a glance what changes we made to the code.
:::

Then we can update our R code so that we can compare the historic and projected trend in fertility rates between Nigeria and Rwanda.

::: grid
::: g-col-6
```{r, echo=TRUE, eval=FALSE}
library(tidyverse) ; library(ggtext)

df <- read_csv("data/fertility_rate.csv") |> 
  mutate(year = ymd(year, truncated = 2L),
         position = if_else(name == "Nigeria", 0, 0.4))

ggplot(df, aes(x = year, y = value)) + 
  geom_line(aes(colour = name), linewidth = 1, show.legend = FALSE) + 
  geom_text(data = filter(df, year == max(year)),
            aes(label = name, colour = name, vjust = position),
            hjust = -0.1, fontface = "bold",
            show.legend = FALSE) +
  scale_colour_manual(values = c("Nigeria" = "#CC79A7", "Rwanda" = "#009E73")) +
  scale_x_date(expand = expansion(),
               breaks = seq.Date(as.Date("1960-01-01"), as.Date("2100-01-01"), by = "20 years"),
               date_labels = "%Y") + 
  scale_y_continuous(limits = c(0,10),
                     labels = scales::label_number(accuracy = 1, scale_cut = scales::cut_short_scale())) +
  labs(x = NULL, y = "Births per woman",
       title = "Total fertility rate",
       subtitle = "<span style = 'color:#757575;'>Nigeria and Rwanda, 1953 to 2100 (projected)</span>",
       caption = "Source: U.S. Census Bureau, International Database") +
  theme_minimal(base_size = 14) +
  theme(
    plot.margin = margin(0.5,2.5,0.5,0.5,"cm"),
    panel.grid.major.y = element_blank(),
    panel.grid.minor = element_blank(),
    axis.line.x = element_line(colour = "#000000"),
    plot.title = element_text(size = rel(1.2), face = "bold"),
    plot.subtitle = element_markdown(margin = margin(b = 15)),
    plot.caption = element_text(size = rel(0.8), colour = "#707071", margin = margin(t = 20)),
    axis.title.y = element_text(size = rel(0.8), hjust = 0, margin = margin(r = 10))
  ) +
  coord_cartesian(clip = "off")

ggsave("outputs/plot.jpeg", dpi = 300, scale = 1)
```
:::

::: g-col-6
```{r}
library(tidyverse) ; library(ggtext)

df <- read_csv("data/fertility_rate.csv") |> 
  mutate(year = ymd(year, truncated = 2L),
         position = if_else(name == "Nigeria", 0, 0.4))

ggplot(df, aes(x = year, y = value)) + 
  geom_line(aes(colour = name), linewidth = 1, show.legend = FALSE) + 
  geom_text(data = filter(df, year == max(year)),
            aes(label = name, colour = name, vjust = position),
            hjust = -0.1, fontface = "bold",
            show.legend = FALSE) +
  scale_colour_manual(values = c("Nigeria" = "#CC79A7", "Rwanda" = "#009E73")) +
  scale_x_date(expand = expansion(),
               breaks = seq.Date(as.Date("1960-01-01"), as.Date("2100-01-01"), by = "20 years"),
               date_labels = "%Y") + 
  scale_y_continuous(limits = c(0,10),
                     labels = scales::label_number(accuracy = 1, scale_cut = scales::cut_short_scale())) +
  labs(x = NULL, y = "Births per woman",
       title = "Total fertility rate",
       subtitle = "<span style = 'color:#757575;'>Nigeria and Rwanda, 1953 to 2100 (projected)</span>",
       caption = "Source: U.S. Census Bureau, International Database") +
  theme_minimal(base_size = 14) +
  theme(
    plot.margin = margin(0.5,2.5,0.5,0.5,"cm"),
    panel.grid.major.y = element_blank(),
    panel.grid.minor = element_blank(),
    axis.line.x = element_line(colour = "#000000"),
    plot.title = element_text(size = rel(1.2), face = "bold"),
    plot.subtitle = element_markdown(margin = margin(b = 15)),
    plot.caption = element_text(size = rel(0.8), colour = "#707071", margin = margin(t = 20)),
    axis.title.y = element_text(size = rel(0.8), hjust = 0, margin = margin(r = 10))
  ) +
  coord_cartesian(clip = "off")
```
:::
:::

To track these changes we'd follow these steps.

1. Check that you have file changes: `git status`{.bash}

```sh
$ git status
On branch master
Changes not staged for commit:
  (use "git add <file>..." to update what will be committed)
  (use "git restore <file>..." to discard changes in working directory)
        modified:   data/fertility_rate.csv
        modified:   outputs/plot.jpeg
        modified:   script.R

no changes added to commit (use "git add" and/or "git commit -a")
```

We have updated our dataset with values for Rwanda (`fertility_rate.csv`), created a new plot (`plot.jpeg`) and revised our R script (`script.R`).

2. Put the changes in the staging area: `git add -A`{.bash}
3. Commit the changes: `git commit -m "Compare with Rwanda"`{.bash}
4. Check that everything has been committed: `git status`{.bash}

::: {.callout-tip}
## `.gitignore`

A `.gitignore` is a text file that lists all of the files and folders that you don't want tracked by Git. For example, this list could include datasets that contain sensitive data. Listing files to `.gitignore` is especially important if you intend to host files on GitHub. To create a `.gitignore` file just enter `touch .gitignore`{.bash} into the terminal. You can either edit the file in the terminal or in a text editor. If you want to add a file using the terminal you can enter `echo FILENAME >> .gitignore`{.bash}.
:::

Rather than overwriting your file with different file names you now have a linear history of commits for the same file. To visualise the history of your commits you can type `git log --reverse --pretty=format:'| %H | %s |'` into the Terminal. This prints out the commit **hash**, a unique identifier for each commit, alongside the commit messages.

```sh
$ git log --reverse --pretty=format:'| %H | %s |'
| b7ab60ae6fae9c6263ff36fd4fcda4f8be874683 | Nigeria fertility rate |
| d2faa1a7784fad0524dc33a8585f8cf7c8340b9c | Compare with Rwanda |
```

The glossary below lists some of the most common terms and Git commands.

|                    |                                         |
|--------------------|:----------------------------------------|
| **repo**           | project folder                          |
| **remote**         | a repo available to all team members    |
| **clone**          | get repo from remote                    |
| **stage**          | prepare files for commit                |
| **commit**         |  take a snapshot of a file or files     |
| **hash**           | commit id                               |
| **pull**           | get commits from a repo                 |
| **push**           | send commits to a repo                  |
| **branch**         | sandbox in a repo                       |
| **main**           | default branch where live code sits     |
| **merge**          | combine all commits on branches         |
| **merge conflict** | conflict contradictory file changes     |
| **pull**           | request propose changes to remote repo  |

: Common terms and Git commands {.hover tbl-colwidths="[20,80]"}

## GitHub
[GitHub](https://github.com/) is a website and service that hosts Git repositories, where developers can store, share, and collaborate on their projects.

Before you can add your local **repo** to GitHub you need to create a **remote** repo on GitHub.

1. Create a new [**repo**](https://github.com/new) called `vital_statistics`. There is no need to add a README file, licence or a `.gitignore` file at this stage. Click through to 'Create repository'.
2. Navigate to '**Quick setup — if you’ve done this kind of thing before**' on the subsequent page and copy the remote repository URL. It will something look like `https://github.com/USERNAME/vital_statistics.git`
3. Next, open the terminal and ensure that it is pointing to the current folder. To print the name of the current folder just enter `pwd`{.bash}
4. Now enter `git remote add origin REMOTE-URL`{.bash} into the terminal replacing `REMOTE-URL` with the remote repository URL from step 2.
5. To check that the correct URL has been set enter `git remote -v`{.bash}
6. Then push the changes you have made to your local repo to GitHub enter `git push origin master`{.bash} You should now see all your code and subfolders on GitHub.

::: {.callout-tip}
## Branches

If you are collaborating with others it is generally a good idea to create what's called a **branch**. This is essentially a sandbox for rough workings out. The working code is on the 'main' branch but the draft, experimental code is held on a separate branch.

These are the steps that you might follow:

1. Create a branch: `git branch development`{.bash}
2. Switch to the development branch: `git checkout development`{.bash}
3. Push changes to the remote branch: `git push origin development`{.bash}
:::

If you are collaborator on this project and want to copy or **fork** the GitHub repository for you to work on locally you can follow these steps.

1. Set you working folder e.g. `cd /path/to/your/folder`{.bash}
2. **Clone** the remote repo in the current directory by entering `git clone https://github.com/USERNAME/vital_statistics.git`{.bash} This is the same remote repository URL that we used earlier.



